Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2011-07-13 Text/migration.tex
Second submission to second referee report

The model disk, based on an equilibrium model from \citet{pickett2003}, orbits a 1 $M_{\odot}$ star and has a mass $M_d =
0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim
\varpi^{-1/2}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk
at $\varpi \approx$ 32 AU, or about 180 yr. The disk is initially located between radial grid zones 30 and 240 and is
close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable \citep[see ][]{durisen2007}
minimum value of 1.38 at radial grid zone 161 (26.7 AU). The computational grid extends radially to 512 zones to
accommodate expansion of the outer disk once GIs become nonlinear. An outflow boundary condition is enforced at the
upper vertical grid boundary, the outer radial grid boundary, and an inner radial boundary at 2 AU. To seed
nonaxisymmetry, the density distribution is given an initial 0.01 \% random cell-to-cell perturbation.

This Letter presents three simulations. The first, which we call the {\sl fiducial run}, is simulated from time $t = 0$
to 21 ORP without a planet. In the second simulation, which we call the {\sl t = 0 planet run}, a one Jupiter-mass
($M_J$) planet with a roughly circular midplane orbit at 25 AU is included in the initial $t = 0$ ORP equilibrium
disk. This run investigates the effects of an existing planet on the growth of GIs, which could occur if disk the disk
undergoes episodic or localized instabilities. In the third simulation, which we call the {\sl t = 10 planet run}, a 1
$M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10.5$ ORP. We estimate semi-major axes $a$
and eccentricities $e$ from the planet's $\varpi(t)$ by using the maximum and minimum radii, $\varpi_{\rm max}$ and
$\varpi_{\rm min}$, over each complete orbit of the planet, where one orbit is $\Delta\phi=2\pi$. Specifically, $a =
(\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm
  min})$. These parameters are intended to highlight the strong orbit-to-orbit variations in the motion of the planet.
A more refined estimate of the orbital elements is unnecessary because the potential of the disk, even if were
axisymmetric, will cause departures from Keplerian dynamics.

@@

{\it The $t = 0$ Planet Run.} Figure \ref{fig:Am} shows that inclusion of even this modest mass planet ($0.7 \%$ of
$M_d$) has a dramatic effect on the burst phase. Without a planet, when all modes grow from imposed noise, it takes 4
ORPs for coherent spiral modes to organize and grow to nonlinear amplitude. The growth is centered near the $Q$-minimum
at $26$ AU. Modes of cos($m\phi$) symmetry with $m = 4$ and 5 dominate. An $m = 3$ (three-armed) spiral also grows, but
somewhat more slowly, and lags the other modes in time. In the $t = 0$ planet run, the planet develops an organized wake
within about an ORP in which $m = 3$ is a dominant component ($\sim 5$ \% global density perturbation). This nonlinear
seeding causes $m = 3$ to dominate the GI burst, which also occurs $\sim$ 2 ORP earlier. Because of our initial
placement of the planet inside but close to the $Q$-minimum, the corotation radius (CR) of the $m = 3$ mode is fairly
close to that of the planet's orbit radius. When the triggered $m = 3$ mode becomes strongly nonlinear at about 3 ORP,
planet migration is significantly affected.

>>>>

The model disk, based on an equilibrium model from \citet{pickett2003}, orbits a 1 $M_{\odot}$ star and has a mass $M_d =
0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim
\varpi^{-1/2}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk
at $\varpi \approx$ 32 AU, or about 180 yr. The disk is initially located between radial grid zones 30 and 240 and is
close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable \citep[see][]{durisen2007}
minimum value of 1.38 at radial grid zone 161 (26.7 AU). The computational grid extends radially to 512 zones to
accommodate expansion of the outer disk once GIs become nonlinear. An outflow boundary condition is enforced at the
upper vertical grid boundary, the outer radial grid boundary, and an inner radial boundary at 2 AU. To seed
nonaxisymmetry, the density distribution is given an initial 0.01\% random cell-to-cell perturbation.

This Letter presents three simulations. The first, which we call the {\sl fiducial run}, is simulated from time $t = 0$
to 21 ORP without a planet. In the second simulation, which we call the {\sl t = 0 planet run}, a one Jupiter-mass
($M_J$) planet with a roughly circular midplane orbit at 25 AU is included in the initial $t = 0$ ORP equilibrium
disk. This run investigates the effects of an existing planet on the growth of GIs, which could occur if disk the disk
undergoes episodic or localized instabilities. In the third simulation, which we call the {\sl t = 10 planet run}, a 1
$M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10.5$ ORP. We estimate semi-major axes $a$
and eccentricities $e$ from the planet's $\varpi_p(t)$ by using the maximum and minimum radii, $\varpi_{\rm max}$ and
$\varpi_{\rm min}$, over each complete orbit of the planet, where one orbit is $\Delta\phi=2\pi$. Specifically, $a =
(\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm
  min})$. These parameters are intended to highlight the strong orbit-to-orbit variations in the motion of the planet.
A more refined estimate of the orbital elements is unnecessary because the potential of the disk, even if were
axisymmetric, will cause departures from Keplerian dynamics.

@@

{\it The $t = 0$ Planet Run.} Figure \ref{fig:Am} shows that inclusion of even this modest mass planet ($0.7 \%$ of
$M_d$) has a dramatic effect on the burst phase. Without a planet, when all modes grow from imposed noise, it takes 4
ORPs for coherent spiral modes to organize and grow to nonlinear amplitude. The growth is centered near the $Q$-minimum
at $26$ AU. Modes of cos($m\phi$) symmetry with $m = 4$ and 5 dominate. An $m = 3$ (three-armed) spiral also grows, but
somewhat more slowly, and lags the other modes in time. In the $t = 0$ planet run, the planet develops an organized wake
within about an ORP in which $m = 3$ is a dominant component ($\sim 5$\% global density perturbation). This nonlinear
seeding causes $m = 3$ to dominate the GI burst, which also occurs $\sim$ 2 ORP earlier. Because of our initial
placement of the planet inside but close to the $Q$-minimum, the corotation radius (CR) of the $m = 3$ mode is fairly
close to that of the planet's orbit radius. When the triggered $m = 3$ mode becomes strongly nonlinear at about 3 ORP,
planet migration is significantly affected.
